TSELM: Target speaker extraction using discrete tokens and language models

### 2024-09-06 src/paper.tex
update

\begin{abstract}
We present TSELM, a novel target speaker extraction network using discrete tokens and language models. This model utilizes multiple discretized layers
of WavLM Large as input tokens. It uses cross attention to inject target speaker information and language models to model the sequences. It uses 
a scalable HiFi-GAN to reconstruct the audio from the tokens. It uses 
cross entropy loss to model the probability of output tokens, transforming the complex regression problems into classification problems. To the best of our 
knowledge, we are the first to do target speaker extraction using discrete tokens. However, there is still a gap between our model and continuous methods. 
Our codes and pretrained models are available at: \href{https://github.com/Beilong-Tang/TSELM}{https://github.com/Beilong-Tang/TSELM} .
\end{abstract}

\begin{IEEEkeywords}
target speaker extraction, speech separation, language models, audio discretization
\end{IEEEkeywords}
\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{assets/model.pdf}
    \caption{Overview of TSELM.}
    \label{model}
    \end{figure*}

@@

Discretization methods have been studied for speech 
enhancement \cite{selm} and blind speech separation \cite{dasb,tokensplit}, 
however, this approach has 
been rarely studied for target speaker extraction. In this paper, we present a novel way to do 
target speaker extraction using discrete tokens and LMs (TSELM). Inspired by the blind speech separation 
networks proposed in DASB \cite{dasb}, our model has three stages: encoding, modeling and decoding.
For the encoding stage, reference speech and mixed speech are tokenized using WavLM and 
Kmeans as tokenizer. Unlike SELM \cite{selm}, which uses the 6th layer of WavLM as input to the Kmeans layer, 
we follow the recipes in \cite{dasb} and uses the output from hidden 
layers 1, 3, 7, 12, 18, 23 of WavLM as input to the Kmeans layer. 6 Kmeans models
are trained and applied on each layer output respectively. 
Our results indicate that using multiple layers as 
input is better 
than using only the 6th layer. The reference speech is fed to the encoder directly. 
However, for the mixed speech, instead of directly passing it to the WavLM model, we first concatenated the reference speech on both sides 
of it before passing. After tokenization, we use only the tokens 
corresponding to the mixed speech. For the modeling stage, we utilized an attention 
embedding mechanism to add each embeddings of all layers together. We used a Cross 
Attention mechanism similar in \cite{sef_net} to inject the speaker information. 
An encoder-only LM, following by a linear classifier, is applied after the cross attention module
to output the reconstructed tokens. 
For the decoding stage, we used the pretrained unit HiFi-GAN 
in \cite{unit_hifi} to reconstruct the 
discrete tokens back to audio. Unlike SELM \cite{selm}, where a conformer 
detokenizer is trained 
to reconstruct the WavLM embeddings using Kmeans center embeddings before HiFi-GAN, \cite{unit_hifi} has proposed a scalable HiFi-GAN using dropout mechanism to directly 
reconstruct audio from multiple layers of tokens without a conformer detokenizer. It 
has also eliminated the trouble of training a HiFi-GAN for each layer. The encoder and 
decoder is freezed during the training. The model overview is in Fig.\ref{model}. 
Through extensive studies, we have demonstrated this method has achieved comparable results in terms 
of speech quality and intelligibility. To the best of our knowledge, we are the first to explore 
using discrete tokens to conduct target speaker extraction. Our demos are available at \href{https://beilong-tang.github.io/TSELM.demo/}{https://beilong-tang.github.io/TSELM.demo/}.

>>>>

\begin{abstract}
  We propose TSELM, a novel target speaker extraction network that leverages discrete tokens and language models. The model utilizes multiple discretized layers from WavLM Large as input tokens and incorporates cross-attention mechanisms to integrate target speaker information. Language models are employed to capture the sequence dependencies, while a scalable HiFi-GAN is used to reconstruct the audio from the tokens. By applying a cross-entropy loss, TSELM models the probability distribution of output tokens, thus converting the complex regression problem of audio generation into a classification task. To the best of our knowledge, this is the first approach to target speaker extraction using discrete tokens. However, a performance gap remains between our method and continuous approaches. The code and pretrained models are available at: \href{https://github.com/Beilong-Tang/TSELM}{https://github.com/Beilong-Tang/TSELM}.
\end{abstract}

\begin{IEEEkeywords}
target speaker extraction, speech separation, language models, audio discretization, Self-Supervised Models
\end{IEEEkeywords}
\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{assets/model.pdf}
    \caption{Overview of TSELM.}
    \label{model}
    \end{figure*}

@@

Discretization methods have been studied for speech 
enhancement \cite{selm} and blind speech separation \cite{dasb,tokensplit}, 
however, it has 
rarely been studied for TSE. In this paper, we present a novel way to do 
target speaker extraction using discrete tokens and LMs called \textbf{TSELM}. Inspired by the blind speech separation 
networks proposed in DASB \cite{dasb}, our model has three stages: encoding, modeling and decoding.
For the encoding stage, reference speech and mixed speech are tokenized using WavLM and 
Kmeans as tokenizer. Unlike SELM \cite{selm}, which uses the 6th layer of WavLM as input to the Kmeans layer, 
we follow the recipes in \cite{dasb} and uses the output from hidden 
layers 1, 3, 7, 12, 18, 23 of WavLM as input to the Kmeans layer. 6 Kmeans models
are trained and applied on each layer output respectively. 
Our results indicate that using multiple layers as 
input is better 
than using only the 6th layer. The reference speech is fed to the encoder directly. 
However, for the mixed speech, instead of directly passing it to the WavLM model, we first concatenated the reference speech on both sides 
of it before passing. After tokenization, we use only the tokens 
corresponding to the mixed speech. For the modeling stage, we utilized an attention 
embedding mechanism to add each embeddings of all layers together. We used a Cross 
Attention mechanism similar in \cite{sef_net} to inject the speaker information. 
An encoder-only LM, following by a linear classifier, is applied after the cross attention module
to output the reconstructed tokens. 
For the decoding stage, we used the pretrained unit HiFi-GAN 
in \cite{unit_hifi} to reconstruct the 
discrete tokens back to audio. Unlike SELM \cite{selm}, where a conformer 
detokenizer is trained 
to reconstruct the WavLM embeddings using Kmeans center embeddings before HiFi-GAN, \cite{unit_hifi} has proposed a scalable HiFi-GAN using dropout mechanism to directly 
reconstruct audio from multiple layers of tokens without a conformer detokenizer. It 
has also eliminated the trouble of training a HiFi-GAN for each layer. The encoder and 
decoder is freezed during the training. The model overview is in Fig.\ref{model}. 
Through extensive studies, we have demonstrated this method has achieved comparable results in terms 
of speech quality and intelligibility. To the best of our knowledge, we are the first to explore 
using discrete tokens to conduct target speaker extraction. Our demos are available at \href{https://beilong-tang.github.io/TSELM.demo/}{https://beilong-tang.github.io/TSELM.demo/}.
